---
title: "Hello R"
author: "Taavi Päll"
date: "26 3 2018"
output: 
  github_document: default
  html_document:
    theme: united
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

Credit: @jennybc, @datacamp

## R basics and Rstudio

- Launch RStudio
- Notice the default panes:
    - This document (upper left)
    - Console (lower left)
    - Environment/History (tabbed in upper right) 
    - Files/Plots/Packages/Help (tabbed in lower right)

## R as calculator

In its most basic form, R can be used as a simple calculator. 
Consider the following arithmetic operators:

- Addition: +
- Subtraction: -
- Multiplication: *
- Division: /
- Exponentiation: ^
- Modulo: %%

```{r}
# An addition
5 + 5
```

```{r}
# A subtraction
5 - 5 
```

```{r}
# A multiplication
3 * 5
```

```{r}
# A division
(5 + 5) / 2
```


### Exercise

1. Calculate 2 to the power of 8?
```{r}
# Exponentiation
```

## Variable assignment

A basic concept in (statistical) programming is called a **variable**.

A variable allows you to store a value (e.g. 4) or an object (e.g. a function description) in R. 
You can then later use this variable's name to easily access the value or the object that is stored within this variable.

You can assign a value 4 to a variable `my_var` with the command:

```{r}
my_var <- 4
```

### Excercise

1. Please complete this code by assigning value 42 to variable x (but first you can try to run this code chunk as it is. What is R complaining about?):
```{r}
# Assign the value 42 to x
x <- 

# Print out the value of the variable x
x
```

`Error: object 'x' not found` means that object with name x does not exists.
Have a look at your Environment. 
Do you see object x? 
You can only create objects by assigning them some value(s) or even empty structure.

> Note that `<-` is the assignment operator in R

2. Suppose you want to calculate __body mass index__. 

Body mass index (BMI) is calculated from height in m and weight in kg using following formula:

$$BMI = \frac{height_{kg}}{weight_m^2}$$

You want to store your body weight in a variable with the name my_weight.

Type the following code into chunk: `my_weight <- 99` (replace 99 with your body weight). 

This will assign the value of your body weight to my_apples.

```{r}
# assign your weight to variable my_weight

```

Type: my_weight into chunk below. 
This will print out the value of my_weight:

```{r}
# type: my_weight

```

BMI formula also needs height.

Please create also the variable **my_height** and assign the value of your height in meters to it: 

```{r}
# create variable my_height with your height in meters

```

Next, you can calculate your body mass index. 
Since you have given meaningful names to these values, you can now code this in a clear way `my_weight / my_height^2`

Calculate BMI and have R simply print the result.
```{r}

```

In order to use calculated BMI value in downstream analyses or visualizations you need to assign it to a new variable, let's call it my_bmi.

Assign the result of BMI calculation to a new variable my_bmi.

```{r}

```

> The big advantage of doing calculations with variables is reusability. 

If you just change my_weight to equal 45 instead of 99, or whatever your body weight is, and rerun the script, my_bmi will automatically update as well.

The + operator works with numeric variables in R. 

If you really tried to calculate BMI using "weight" and "height", and assigned a text value to the variable my_weight (see the chunk below), you would be trying to assign the calculation using of a numeric and a character variable to the variable my_height. This is not possible!

```{r}
my_weight <- "forty five"
```

```{r}
my_bmi <- my_weight / (my_height^2)
```

Try to fix this code above.

## Basic data types in R

R works with numerous data types. 
Some of the most basic types to get started are:

- Decimals values like 4.5 are called numerics.

- Natural numbers like 4 are called integers. Integers are also numerics.

- Boolean values (TRUE or FALSE) are called logical.

- Text (or string) values are called characters.

```{r}
# Change my_numeric to be 42
my_numeric <- 42.5

# Change my_character to be "universe"
my_character <- "some text"

# Change my_logical to be FALSE
my_logical <- TRUE
```

Note how the quotation marks on the chunk above indicate that "some text" is a character.

### What's that data type?

When you tried to calculate BMI using my_weight with value "forty five", you got an error due to a mismatch in data types.

You can avoid such annoying situations by checking the data type of a variable beforehand. 

You can do this with the class() function, as the code below shows.

```{r}
# Declare variables of different types
my_numeric <- 42
my_character <- "universe"
my_logical <- FALSE 

# Check class of my_numeric
class(my_numeric)

# Check class of my_character


# Check class of my_logical

```

